---
title: "Matilda in R"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

The complete Matilda workflow in **R** on real **TEA-seq** (RNA + ADT + ATAC): read your data →
train → classification → dimension reduction → feature selection → simulation, plus a look at the
different data types Matilda supports. (Python users: a parallel tutorial with the same structure
is available as a Jupyter notebook.)

## Requirements

- **R**: `matilda`, `SingleCellExperiment`; `Seurat` for the `10x` and `.rds` loaders; `scater`
  for UMAP; `ggplot2` for plots.
- **Python**: none — `matilda` provisions it via `basilisk` on the first `matilda_train()`.
- **Demo data**: downloaded automatically by `matilda_download_example()` (~75 MB, cached locally).

```{r deps}
library(matilda)
library(SingleCellExperiment)
data_dir <- matilda_download_example()         # downloads + caches the TEA-seq demo, returns the path
demo_h5  <- data_dir                            # native .h5 + _cty.csv
demo_fmt <- file.path(data_dir, "formats")      # .h5ad / 10x / .rds

# a fixed colour list, assigned to whichever cell types your data has (works for any labels)
PALETTE <- c("#1f77b4","#aec7e8","#ff7f0e","#ffbb78","#2ca02c","#98df8a","#d62728","#ff9896",
             "#9467bd","#c5b0d5","#8c564b","#c49c94","#e377c2","#f7b6d2","#7f7f7f","#bcbd22","#17becf","#9edae5")
```

# 1. Read your data

Matilda's R interface takes a `SingleCellExperiment` — **RNA raw counts** in the main assay, **ADT**
and **ATAC** (gene activity) as `altExp`s, and a cell-type column in `colData`. Load each modality with
the reader for its format; the demo ships the same training data in several formats, so here we read RNA
from `.h5`, ADT from `.h5ad`, and ATAC from 10x just to show each reader — for your own data you'd
normally use one format for all modalities. (A `Seurat` `.rds` works too — see the note below.)

```{r read-data}
read_h5 <- function(path) {                                   # native Matilda .h5 (rhdf5) -> genes x cells
  m     <- rhdf5::h5read(path, "matrix/data")
  feats <- as.character(rhdf5::h5read(path, "matrix/features"))
  cells <- as.character(rhdf5::h5read(path, "matrix/barcodes"))
  if (nrow(m) == length(cells) && ncol(m) == length(feats)) m <- t(m)
  dimnames(m) <- list(feats, cells); m
}
rd_h5ad <- function(p) {                                      # .h5ad (rhdf5, CSR) -> genes x cells
  X   <- rhdf5::h5read(p, "X"); shp <- as.integer(rhdf5::h5readAttributes(p, "X")$shape)
  m <- Matrix::sparseMatrix(i = rep(seq_len(shp[1]), diff(X$indptr)), j = X$indices + 1L,
                            x = as.numeric(X$data), dims = shp)
  m <- Matrix::t(m)
  dimnames(m) <- list(as.character(rhdf5::h5read(p, "var/_index")),
                      as.character(rhdf5::h5read(p, "obs/_index"))); as.matrix(m)
}
rd_10x <- function(d) as.matrix(Seurat::Read10X(d))           # 10x CellRanger (Seurat) -> genes x cells

build_sce <- function(rna, adt, atac, cell_type) {
  sce <- SingleCellExperiment(assays = list(counts = as.matrix(rna)))
  altExp(sce, "ADT")  <- SummarizedExperiment(list(counts = as.matrix(adt)))
  altExp(sce, "ATAC") <- SummarizedExperiment(list(counts = as.matrix(atac)))
  sce$cell_type <- cell_type; sce
}
labels <- as.character(utils::read.csv(file.path(demo_h5, "train_cty.csv"), header = FALSE)[[2]])[-1]

sce <- build_sce(read_h5(file.path(demo_h5,  "train_rna.h5")),                 # native .h5
                 rd_h5ad(file.path(demo_fmt, "train_adt.h5ad")),              # .h5ad
                 rd_10x(file.path(demo_fmt,  "10x", "train_atac")),           # 10x CellRanger
                 labels)
sce
```

A `Seurat` `.rds` (R-only) works the same way — pull each assay's counts and pass to `build_sce`:
`seu <- readRDS(file.path(demo_fmt, "train_seurat.rds")); SeuratObject::LayerData(seu, assay = "RNA", layer = "counts")`.

# 2. Train the Matilda model

The model is stored inside the object, so the workflow pipes.

```{r train}
sce <- matilda_train(sce, label = "cell_type", seed = 1L)
matilda_model(sce)
```

# 3. Classification (held-out query cells)

```{r classify}
query <- build_sce(read_h5(file.path(demo_h5, "test_rna.h5")),
                   read_h5(file.path(demo_h5, "test_adt.h5")),
                   read_h5(file.path(demo_h5, "test_atac.h5")),
                   as.character(utils::read.csv(file.path(demo_h5, "test_cty.csv"), header = FALSE)[[2]])[-1])
query <- matilda_classify(query, reference = sce)
sprintf("overall query accuracy = %.4f", mean(query$matilda_pred == query$cell_type))
```

```{r classify-plot, fig.width = 7, fig.height = 5}
acc <- tapply(query$matilda_pred == query$cell_type, query$cell_type, mean)
ggplot(data.frame(cell_type = names(acc), accuracy = as.numeric(acc)),
       aes(reorder(cell_type, accuracy), accuracy)) +
  geom_col(fill = "steelblue") + coord_flip() + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.2f", accuracy)), hjust = -0.1, size = 3) +
  labs(title = "Query accuracy by cell type", x = NULL, y = "accuracy") + theme_classic()
```

# 4. Dimension reduction

```{r reduce}
sce <- matilda_reduce(sce)                 # adds reducedDim(sce, "MATILDA")
dim(reducedDim(sce, "MATILDA"))            # cells x z_dim (100)
```

```{r umap, fig.width = 7, fig.height = 5, eval = requireNamespace("scater", quietly = TRUE)}
set.seed(42)
sce <- scater::runUMAP(sce, dimred = "MATILDA", n_neighbors = 15L, min_dist = 0.1)
cats <- sort(unique(as.character(sce$cell_type)))                 # colour whatever cell types the data has
cols <- setNames(PALETTE[(seq_along(cats) - 1) %% length(PALETTE) + 1], cats)
scater::plotUMAP(sce, colour_by = "cell_type") +
  ggplot2::scale_colour_manual(values = cols) +
  ggtitle("Matilda integrated latent space (UMAP)")
```

# 5. Feature selection

```{r markers}
markers <- matilda_markers(sce)            # data.frame: celltype / feature / importance
head(markers)
```

```{r markers-plot, fig.width = 8, fig.height = 6}
# order cell types alphabetically; group features by the cell type they top-3 mark, then importance
ord_ct <- sort(unique(markers$celltype))
top    <- do.call(rbind, lapply(ord_ct, function(ct) {
  d <- markers[markers$celltype == ct, ]; head(d[order(-d$importance), ], 3)
}))
feat_order <- unique(top$feature)
sub <- markers[markers$feature %in% feat_order, ]
sub$imp_s    <- ave(sub$importance, sub$feature, FUN = function(z) (z - min(z)) / (max(z) - min(z) + 1e-9))
sub$feature  <- factor(sub$feature,  levels = rev(feat_order))
sub$celltype <- factor(sub$celltype, levels = ord_ct)
ggplot(sub, aes(celltype, feature, fill = imp_s)) +
  geom_tile() + scale_fill_viridis_c() + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Top-3 markers per cell type", x = NULL, y = NULL, fill = "scaled\nimportance")
```

# 6. Simulation

Matilda augments a chosen **anchor** cell type and returns the synthetic cells. The returned object
also carries the real reference cells Matilda used (`metadata(sim)$real`), in the **same feature
space** — so the two are directly comparable. The official check: per modality, UMAP the real and
simulated cells together, then show the real cells (top row) and the simulated cells (bottom row),
highlighting the anchor. If augmentation works, the anchor lands in the same place in both rows.

```{r simulate}
sim  <- matilda_simulate(sce, celltype = "B.Naive", n = 200)
real <- S4Vectors::metadata(sim)$real        # real reference cells, same space as the simulation
c(simulated = ncol(sim), real = ncol(real$rna), anchor_cells = sum(sim$label == "B.Naive"))
```

```{r sim-plot, fig.width = 9, fig.height = 6.5, eval = requireNamespace("uwot", quietly = TRUE)}
real_mat <- list(RNA = real$rna, ADT = real$adt, ATAC = real$atac)
sim_mat  <- list(RNA  = SummarizedExperiment::assay(sim, "counts"),
                 ADT  = SummarizedExperiment::assay(SingleCellExperiment::altExp(sim, "ADT"),  "counts"),
                 ATAC = SummarizedExperiment::assay(SingleCellExperiment::altExp(sim, "ATAC"), "counts"))
anchor_tag <- function(lab) factor(ifelse(lab == "B.Naive", "anchor cell type", "other cell types"),
                                   levels = c("anchor cell type", "other cell types"))
one_modality <- function(mod) {                # per modality: top-1000 HVG -> scale -> PCA -> UMAP, then split
  X  <- t(cbind(real_mat[[mod]], sim_mat[[mod]]))                 # cells x features
  nr <- ncol(real_mat[[mod]]); ns <- ncol(sim_mat[[mod]])
  hv <- order(apply(X, 2, stats::var), decreasing = TRUE)[seq_len(min(1000L, ncol(X)))]
  Z   <- scale(X[, hv])
  pcs <- stats::prcomp(Z, rank. = min(50L, ncol(Z) - 1L), center = FALSE, scale. = FALSE)$x
  set.seed(42)
  u <- as.data.frame(uwot::umap(pcs, n_neighbors = 20, min_dist = 0.1, init = "pca"))
  names(u) <- c("UMAP1", "UMAP2")
  rbind(data.frame(u[seq_len(nr), ],      grp = anchor_tag(real$label), panel = paste("Real", mod)),
        data.frame(u[nr + seq_len(ns), ], grp = anchor_tag(sim$label),  panel = paste("Simulated", mod)))
}
big <- do.call(rbind, lapply(c("RNA", "ADT", "ATAC"), one_modality))
big$panel <- factor(big$panel, levels = c("Real RNA", "Real ADT", "Real ATAC",
                                          "Simulated RNA", "Simulated ADT", "Simulated ATAC"))
ggplot(big, aes(UMAP1, UMAP2, colour = grp)) +
  geom_point(size = 0.5, alpha = 0.8) +
  facet_wrap(~ panel, nrow = 2, scales = "free") +
  ggplot2::scale_colour_manual(values = c("anchor cell type" = "#F8766D", "other cell types" = "#00BFC4")) +
  theme_classic() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.title = element_blank(), aspect.ratio = 1) +
  labs(x = NULL, y = NULL, title = "Real vs. simulated cells per modality (anchor: B.Naive)")
```

# 7. Modality combinations Matilda supports

Matilda works with **any combination of the three modalities** — RNA only, RNA + ADT, RNA + ATAC,
or RNA + ADT + ATAC — and picks the matching model automatically from what you provide. (Internally
Matilda labels these `rna_only` / `CITEseq` / `SHAREseq` / `TEAseq`, after the assays they typically
come from, but your data needn't be those specific protocols — only the modalities present matter.)
Training each combination on the *same* cells shows what each added modality buys; to use your own
data, just build the SCE with whatever modalities you have.

```{r modality, fig.width = 7, fig.height = 4}
keep_only <- function(s, keep) { for (ae in altExpNames(s)) if (!(ae %in% keep)) altExp(s, ae) <- NULL; s }
acc <- sapply(list("RNA only" = character(0), "RNA+ATAC" = "ATAC", "RNA+ADT" = "ADT"),
  function(keep) {
    fit <- matilda_train(keep_only(sce, keep), label = "cell_type", seed = 1L)
    qy  <- matilda_classify(keep_only(query, keep), reference = fit)
    mean(qy$matilda_pred == qy$cell_type)
  })
acc["RNA+ADT+ATAC"] <- mean(query$matilda_pred == query$cell_type)   # the TEAseq model from §2/§3
ggplot(data.frame(modalities = factor(names(acc), levels = names(acc)), accuracy = as.numeric(acc)),
       aes(modalities, accuracy)) +
  geom_col(fill = "seagreen") + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.3f", accuracy)), vjust = -0.4, size = 3.5) +
  labs(title = "Query accuracy by modality combination", x = NULL, y = "accuracy") + theme_classic()
```

Adding the ADT (protein) panel to RNA gives the biggest gain; ATAC contributes less for this task
on this dataset.

# 8. Session info

```{r sessioninfo}
sessionInfo()
```
